---
title: "easy-downscaling"
author: "Violet Lingenfelter"
date: "4/8/2020"
output: html_document
---
# Downscaling Climate Data Using Very Primitive Interpolation 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(rgbif) # include library for getting species distribution data from GBIF
library(ggplot2) # include library for making maps
library(raster) # include library for reading raster files
library(dplyr)
require(maps)
require(viridis)
library(ncdf4) # package for netcdf manipulation
library(rgdal) # package for geospatial analysis
library(gstat)

theme_set(
  theme_void()
)

setwd('/Users/vlingenfelter5/Documents/ecological-dynamics/final-project/')
```

# Temperature data

For each year we need a raster where each cell has the warmest value predicted for that year (the max value of all months) and the coolest value predicted for that year (the minimum value for that month). 

```{r}
ts_files <- list.files('/Volumes/OBESE/SDS Research/ts/') # get all of the surface temperature files that I have downloaded to an external drive

ts_data <- list()

for (i in 1:length(ts_files)) {
  print(ts_files[i]) # print name of file
  temp <- raster(paste("/Volumes/OBESE/SDS Research/ts/", ts_files[[i]], sep="")) # get data from folder on hard drive
  temp <- rotate(temp) # change raster to normal degree scheme 
  temp <- crop(temp, c(-180,0,0,90)) # crop to the western northern hemispheres
  ts_data[[i]] <- disaggregate(temp, fact=8) # this is the simplest case of downscaling (raster disaggregation) to the 10 arcmin resolution (1/8 of a degree)
}

coldest_cold <- list() # empty list to hold the coldest cold data for each year
warmest_warm <- list() # empty list to hold the warmest warm data for each year

# for every year, get the rasters from that year, then calculate the max value and min value for each cell
for (i in 1:(length(ts_data)/12)) {
  print(2014+i)
  subset <- ts_data[(((i-1)*12)+1):(i*12)]
  warmest_warm[i] <- overlay(subset[[1]], subset[[2]], subset[[3]], subset[[4]], subset[[6]], subset[[7]], subset[[8]], subset[[9]], subset[[10]], subset[[11]], subset[[12]], fun = max)
  coldest_cold[i] <- overlay(subset[[1]], subset[[2]], subset[[3]], subset[[4]], subset[[6]], subset[[7]], subset[[8]], subset[[9]], subset[[10]], subset[[11]], subset[[12]], fun = min)
}

# write these raster files to disk
for (i in 1:length(coldest_cold)) {
  writeRaster(coldest_cold[[i]], paste("coldest_cold", toString(i+2014), sep="_"), format="CDF")
  writeRaster(warmest_warm[[i]], paste("warmest_warm", toString(i+2014), sep="_"), format="CDF")
}
```

# Precipitation data

For each year we need a raster where each cell has the wettest value predicted for that year (the max value of all months) and the driest value predicted for that year (the minimum value for that month). 

```{r}
pr_files <- list.files('/Volumes/OBESE/SDS Research/pr/') # get all of the precipitation flux files that I have downloaded to an external drive

pr_data <- list() # intialize empty list to store data

for (i in 1:length(pr_files)) {
  print(pr_files[i]) # print name of file
  temp <- raster(paste("/Volumes/OBESE/SDS Research/ts/", ts_files[[i]], sep="")) # get data from folder on hard drive
  temp <- rotate(temp) # change raster to normal degree scheme 
  temp <- crop(temp, c(-180,0,0,90)) # crop to the western northern hemispheres
  pr_data[[i]] <- disaggregate(temp, fact=8) # this is the simplest case of downscaling (raster disaggregation) to the 10 arcmin resolution (1/8 of a degree)
}

driest_dry <- list() # empty list for minimum values
wettest_wet <- list() # empty list for maximum values

# for each year, select all rasters and then find the maximum and minimum cell values
for (i in 1:(length(pr_data)/12)) {
  print(2014+i)
  subset <- pr_data[(((i-1)*12)+1):(i*12)]
  wettest_wet[i] <- overlay(subset[[1]], subset[[2]], subset[[3]], subset[[4]], subset[[6]], subset[[7]], subset[[8]], subset[[9]], subset[[10]], subset[[11]], subset[[12]], fun = max)
  driest_dry[i] <- overlay(subset[[1]], subset[[2]], subset[[3]], subset[[4]], subset[[6]], subset[[7]], subset[[8]], subset[[9]], subset[[10]], subset[[11]], subset[[12]], fun = min)
}

# write these rasters as files to the hard drive
for (i in 1:length(wettest_wet)) {
  writeRaster(wettest_wet[[i]], paste("wettest_wet", toString(i+2014), sep="_"), format="CDF")
  writeRaster(driest_dry[[i]], paste("driest_dry", toString(i+2014), sep="_"), format="CDF")
}
```
